Binning Strategies and Related Loss for Binned Scatterplots of Large Data

### 2016-12-04 AuthorResponseToReviewers.Rmd

*I don’t buy the advice to “use between four and seven frequency”. There simply isn’t empirical evidence in the paper to support this claim.*

- We do not want to allow more color bin than we can *easily* differentiate because the our expressed purpose for working with discretized color scales in this scenario is to have made loss of frequency information *entirely* explicit. While working with discretizing shades we found ourselves worrying about our ability to perfectly match bins to a legend after approximately 6 shades. We do not claim to have gathered empirical evidence, rather we defer to Healey and Enn's Theory on discernible colors (typically 4-7) to back out experience. Figure 9 was included for the reader to assess if they felt our suggestion was justifiable. A statement clarifying that the recommendations is to bolster the reliable perceptual mapping has been added to the discussion section.

*From a human-computer interaction perspective, it is also limiting when the paper talks about spatial loss and frequency loss without any mention of user tasks (e.g. comparison, outlier detection). Do these metric matter for all the tasks involved in reading the scatterplots?*

*Summary*
*This paper feels like a survey paper but the references are not properly cited. As a research paper, it is not clear to me what are the main contributions, and how these contributions build on top of prior work (given that there is no “related work” section in the paper). While the two loss metrics seem potentially interesting, I think more empirical studies are needed with real human participants. It will also improve the paper if the authors can provide take-away guidelines or algorithms that can be applied to real-world problems.*

- The main result from the paper is to provide simple adjustment recommendations for binning specification for reducing loss in binned scatterplots. To emphasize these recommendations, the baseball plot example in Section 5.3 has been reworked to highlight the take-home message. To your other point, we agree that future work to quantify the implicit loss of information (e.g. the imprecise ability to map continuously shaded tiles to a frequency legend) in reading binned scatterplots is needed; however, the current work strives to quantify explicit loss of information when the perception tasks are more simple (e.g. mapping distinguishably-discrete shaded tiles to an interval on the frequency legend).

### Responses to Editors Comments/Questions
*Editor's Comments to Author:*

*Associate Editor - Comments to the Author:*
*The paper in its current form does not provide a sufficiently thorough coverage of either review material or novel material to warrant publication.  However, the reviewers have some useful suggestions for ways in which a different paper (or two) could possibly be produced by expanding on different parts of this submission.*

### 2016-12-04 versionForJCGS/PaperJCGS.Rnw
Address reviewer 2's summary remarks

For the univariate case with observations, $x_i$ for $i \in \{1,\dots,n\}$, binning algorithms require a set of bin centers $x_j^\ast$ for $j \in \{1,\dots,J\}$ and a binning function $b_X(.) : x_i \rightarrow x^\ast_j$ that maps observations to  bin centers. \ktm{What we will refer to as} \textit{General rectangular binning} accomplishes this by defining a sequence of $J$ adjacent intervals, $(\beta_{j-1},\beta_{j}]$ for $j \in \{1,\dots,J\}$, which span over the range of the data. Note that half open intervals are used such that any observation falling on a bin boundary is assigned to a unique interval. Values $x_i$ exactly equal to the lowest bin boundary $\beta_0$ are grouped into the first bin to close the leftmost bound. Each observation is then mapped to a bin center, $x_j^\ast$; the midpoint for the interval to which the observation belongs.

@@

An alternative to the rectangular binning process, is a \textit{random binning} algorithm which utilizes a non-deterministic bin function $b^r_X(\cdot)$ to randomly assigns an observation, $x_i$, to a bin center, $x^\ast$, from a set of possible bins. In this paper, we will consider the simplest case of just two bins, so that without loss of generality we can assume that $x_i$ lies between bin centers $x^\ast_j$ and $x^\ast_{j+1}$. The bin function assigns $x_i$ to a bin center with a probability inversely proportional to the distance to that bin center; the closer a value is to a bin center, the higher the probability the value is assigned to that bin center. More formally,
%
\begin{eqnarray}\label{randbin1}
b^r_X(x_i) = \left\{\begin{array}{ll} 
  x^\ast_j & \text{ with probability } (x^\ast_{j+1} - x_i)/(x^\ast_{j+1} -x^\ast_{j}) \\
  x^\ast_{j+1} &\text{ with probability }  (x_i - x^\ast_{j})/(x^\ast_{j+1} -x^\ast_{j})
  \end{array}\right.
\end{eqnarray}  
%
for $x_i \in [x^\ast_{j+1}, x^\ast_{j}]$. In Table~\ref{tab:rectbinning} we note that this random binning algorithm does not specify bin boundaries; only a sequence of bin centers. This method is easily extensible to also map $x_i$ into more than two bins and can accommodate non-uniform distribution of bin centers. \ktm{The impetus for developing the random binning assignment was to soften the hard breaks associated with one-sided interval assignment for scenarios where many points fall directly on bin boundaries by allowing points to be divided stochastically.} The bin boundaries and centers for standard and random rectangular binning algorithms can be found in Table~\ref{tab:rectbinning}.

>>>>

For the univariate case with observations, $x_i$ for $i \in \{1,\dots,n\}$, binning algorithms require a set of bin centers $x_j^\ast$ for $j \in \{1,\dots,J\}$ and a binning function $b_X(.) : x_i \rightarrow x^\ast_j$ that maps observations to  bin centers. \ktm{What we will refer to as} \textit{general rectangular binning} accomplishes this by defining a sequence of $J$ adjacent intervals, $(\beta_{j-1},\beta_{j}]$ for $j \in \{1,\dots,J\}$, which span over the range of the data. Note that half open intervals are used such that any observation falling on a bin boundary is assigned to a unique interval. Values $x_i$ exactly equal to the lowest bin boundary $\beta_0$ are grouped into the first bin to close the leftmost bound. Each observation is then mapped to a bin center, $x_j^\ast$; the midpoint for the interval to which the observation belongs.

@@

As an alternative to the rectangular binning process, \ktm{we propose} a \textit{random binning} algorithm which utilizes a non-deterministic bin function $b^r_X(\cdot)$ to randomly assigns an observation, $x_i$, to a bin center, $x^\ast$, from a set of possible bins. In this paper, we will consider the simplest case of just two bins, so that without loss of generality we can assume that $x_i$ lies between bin centers $x^\ast_j$ and $x^\ast_{j+1}$. The bin function assigns $x_i$ to a bin center with a probability inversely proportional to the distance to that bin center; the closer a value is to a bin center, the higher the probability the value is assigned to that bin center. More formally,
%
\begin{eqnarray}\label{randbin1}
b^r_X(x_i) = \left\{\begin{array}{ll} 
  x^\ast_j & \text{ with probability } (x^\ast_{j+1} - x_i)/(x^\ast_{j+1} -x^\ast_{j}) \\
  x^\ast_{j+1} &\text{ with probability }  (x_i - x^\ast_{j})/(x^\ast_{j+1} -x^\ast_{j})
  \end{array}\right.
\end{eqnarray}  
%
for $x_i \in [x^\ast_{j+1}, x^\ast_{j}]$. In Table~\ref{tab:rectbinning} we note that this random binning algorithm does not specify bin boundaries; only a sequence of bin centers. This method is easily extensible to also map $x_i$ into more than two bins and can accommodate non-uniform distribution of bin centers. \ktm{The impetus for developing the random binning assignment was to soften the hard breaks associated with one-sided interval assignment for scenarios where many points fall directly on bin boundaries by allowing points to be divided stochastically.} The bin boundaries and centers for standard and random rectangular binning algorithms can be found in Table~\ref{tab:rectbinning}.
